The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2016-03-25 Papers/2014/Universalis/ModalCollapse.tex
Address reviewer's criticisms and Christoph's marginpars

%----------classification, keywords, date
\subjclass{
Prim. 03A02;  % Philosophical aspects of logic and foundations
Sec. 68T02 % Artificial Intelligence
\marginpar{Change?: 03B15, 03B45, 03B35, 68T15}
}

@@

Nevertheless, G{\"o}del's work still leaves room for criticism.  In
particular, his axioms are so strong that, when assuming unrestricted
comprehension principles\footnote{A possible direction to remedy modal
  collapse, as studied e.g. by Koons
  \cite{koons06:_sobel_goedel_ontol_proof} is to impose restrictions
  on the domain of properties.}, 
\marginpar{added footnote}
they entail a \emph{modal collapse}
\cite{Sobel1987,sobel2004logic}: everything that is the case is so
necessarily.  There has been an impressive body of recent and ongoing
work
(cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,Hajek2002,Hajek2008,ContemporaryBibliography}
and the references therein) proposing solutions for the modal
collapse.  The goal of this contribution is to discuss the modal
collapse from the point of view of the modal square of opposition.
\marginpar{@Bruno: Shall we add a sentence or two here?}

@@

Independently of the variant of the ontological argument 
under consideration, the following can be said about 
classes of collapsing propositions:
\begin{enumerate}
\item Valid propositions are collapsing: 
if $\phi$ is valid, then $\collapse{\phi}$ is valid.
%
\item The class of collapsing propositions 
is closed under logical equivalence: 
if $\collapse{\phi}$ is valid and $\phi \biimp \phi'$ is valid, 
then $\collapse{\phi'}$ is valid.
%
\item The class of collapsing propositions is 
not generally closed under equi-validity: 
even if $\collapse{\phi}$ is valid and $\phi$ and $\phi'$ 
are equi-valid, \\ $\collapse{\phi'}$ may not be valid.
%
\item The class of collapsing propositions is 
not generally closed under implication: 
even if $\collapse{\phi}$ is valid and 
$\phi \imp \phi'$ is valid, 
$\collapse{\phi'}$ may not be valid.
\end{enumerate} \marginpar{@Bruno: Shall we respond here to the
  rewievers criticism?}

>>>>

%----------classification, keywords, date
\subjclass{
Prim. 03B15; % Higher-order logic and type theory
Sec. 68T15 % Theorem proving (deduction, resolution, etc.)
}
% \marginpar{Change?: 
% 03A02 % Philosophical aspects of logic and foundations
% 03B45 % Modal logic (including the logic of norms)
% 03B35 % Mechanization of proofs and logical operations
% 68T02 % Artificial Intelligence
% }

@@

Nevertheless, G{\"o}del's work still leaves room for criticism.  In
particular, his axioms are so strong that, when assuming unrestricted
comprehension principles\footnote{A possible direction to remedy modal
  collapse, as studied e.g. by Koons
  \cite{koons06:_sobel_goedel_ontol_proof} is to impose restrictions
  on the domain of properties.}, 
%\marginpar{added footnote}
they entail a \emph{modal collapse}
\cite{Sobel1987,sobel2004logic}: everything that is the case is so
necessarily.  There has been an impressive body of recent and ongoing
work
(cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,Hajek2002,Hajek2008,ContemporaryBibliography}
and the references therein) proposing solutions for the modal
collapse.  The goal of this contribution is to discuss the modal
collapse from the point of view of the modal square of opposition. 
Ontological arguments typically rely on an inversion of the normal direction of entailment
in the modal square of opposition for one particular proposition (i.e. God's existence), 
and the modal collapse shows that this inversion happens for all propositions, resulting in a total collapse of the modal square of opposition.

%\marginpar{@Bruno: Shall we add a sentence or two here?}

@@

In any modal logic at least as strong as \textbf{K}, and even without relying on axioms specific to 
ontological arguments, it is easy to see (and even easier to check with an automated theorem prover) 
the following facts about classes of collapsing propositions:
\begin{enumerate}
\item Valid propositions are collapsing: 
if $\phi$ is valid, then $\collapse{\phi}$ is valid.
%
\item The class of collapsing propositions 
is closed under logical equivalence: 
if $\collapse{\phi}$ is valid and $\phi \biimp \phi'$ is valid, 
then $\collapse{\phi'}$ is valid.
%
\item The class of collapsing propositions is 
not generally closed under equi-validity: 
even if $\collapse{\phi}$ is valid and $\phi$ and $\phi'$ 
are equi-valid, \\ $\collapse{\phi'}$ may not be valid.
%
\item The class of collapsing propositions is 
not generally closed under implication: 
even if $\collapse{\phi}$ is valid and 
$\phi \imp \phi'$ is valid, 
$\collapse{\phi'}$ may not be valid.
\end{enumerate}

An easy corollary of the second fact above is that any ontological argument relying on Lemma \textbf{L2} will necessarily lead to a modal collapse for all propositions that are logically equivalent to God's existence. The third and fourth facts above indicate that to prove that the modal collapse holds for a larger class of propositions would require using axioms specific to the ontological argument.

% \marginpar{@Bruno: Shall we respond here to the
%  rewievers criticism?}
